---
title: 'Stat 439: Homework 2'
subtitle: 'Due Thur 2/10/22 by 5pm in D2L'
author: 'Your name here'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(mosaic)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```


## Instructions

You are strongly encouraged to use R Markdown to complete your Homework assignments, starting with this file as a template. If using R for your answer, include all relevant R code and output. Your homework submission in D2L should be a single pdf file (do not turn in the .Rmd file).

## Part I: Book Exercises

Agresti Exercises 2.8, 2.30ab

## Part II

### 1.

A study was conducted to test the claim that the intake of one
gram of vitamin C per day substantially reduces the frequency
of colds. Volunteers were sought from a variety of occupations
and age groups. Each subject was randomly assigned to either
a 250 mg Vitamin C tablet or a placebo tablet per day for
one month. At the end of the month, researchers recorded whether or not each subject developed a cold. Of the 407 subjects
assigned to vitamin C, 302 developed a cold; of the 411 subjects assigned to the placebo, 335 developed a cold.

a. Assume the sample size was fixed by the researchers. What type of sampling was used in this study: "binomial" or "multinomial." Justify your answer.
a. Is this a randomized experiment or an observational study? Justify your answer.
a. Enter these data into R as a 2x2 table and display the output. The response variable should be entered as the columns, with a "success" as column 1; the explanatory variable should be entered as rows with the base line group as column 1. (This will make the use of the `OddsRatio` function easier.)
a. Create an appropriate plot that displays the relationship between whether the subject took vitamin C or the placebo and whether the subject developed a cold or not. Write one or two sentences describing what the plot tells you.
a. Calculate the estimated relative risk of developing a cold for the vitamin C group compared to the placebo. Fill in the blank in the following interpretation:  "The risk of developing a cold for the vitamin C group was _____% less than the risk of developing a cold in the placebo group."
a. Calculate the estimated relative risk of developing a cold for the placebo group compared to the vitamin C group. Fill in the blank in the following interpretation:  "The risk of developing a cold for the placebo group was _____% less than the risk of developing a cold in the vitamin C group." Explain mathematically why the value in the blank does not equal the value in the blank from part c.
a. Calculate the estimated odds ratio with the vitamin C group in the numerator. Find and interpret an approximate 95% confidence interval for this odds ratio. (Calculations should be done 
by hand, though you may check your confidence interval calculation using the `oddsRatio` function.)


### 2.

Psychologists performed a randomized experiment on 50 male bank supervisors attending a management institute, to investigate biases against women in personnel decisions. The supervisors were asked to make a decision on whether to promote a hypothetical applicant based on a personnel file. For 26 of them, the application file described a female candidate; for the others it described a male. The files were identical in all other respects. Results on the promotion decisions for the two groups are shown below.

```{r}
bank_table <- as.table(
                 matrix(c(21, 3, 14, 12),
                     nrow = 2, ncol = 2,
                     byrow = TRUE,
                     dimnames = list(
                       applicant = c("male", "female"),
                       promoted = c("yes","no")
                       )
                     )
              )
bank_table
```

a. Carry out a randomization test for these data using the odds ratio for the statistic. Include: (i) definition of the parameter, (ii) hypotheses, (iii) value of the estimated odds ratio, (iv) p-value, and (v) conclusion. Your conclusion should also include the scope of inference for this study; that is, to which population can you generalize? and can you conclude causation?
a. Find the p-value for the Fisher's Exact Test of these data using only the `dhyper` and/or `phyper` functions in R (though you may check your work with the `fisher.test` function). Does the p-value from this exact test match the p-value for the randomization test?

