---
title: 'Stat 439: Homework 4'
author: 'Your name here'
subtitle: 'Due Thur 3/10/22 by 11pm in Gradescope'
urlcolor: 'blue'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(alr4) # You may have to install this package
```
## Instructions

You are strongly encouraged to use R Markdown to complete your Homework assignments, starting with this file as a template and Knitting to pdf. Submit your homework to [Gradescope](https://www.gradescope.com/) as a single pdf file.

## Part I: Donner Party

For this problem, we are going to use the built-in `Donner`
data set in the `alr4` library. Examine the help file for a
description of the data.
```{r, echo=FALSE}
?Donner
```
Read the data set into your R session:
```{r}
data(Donner)
```
Note that R reads in all variables except `age` as a factor.
The `glm` function requires that your binary response be coded
as 0's and 1's, so first create an indicator variable for survival:
```{r}
Donner$survive <- as.numeric(Donner$y == "survived")
```

## Q1
How many people in the Donner Party survived? What proportion survived?

## Q2
Fit a logistic regression model to these data using only `age` as the predictor,
and call this model `mod1`.
Report a summary of the model. Then answer the following:

a. Write out the equation of the fitted model. Define any symbols used.
b. Create a scatterplot with `age` on the x-axis and `survive` on the y-axis.
Add a nonparametric smoother (e.g., `lowess`), and add the fitted model curve.
Ensure your plot is well-labeled.
c. Write a sentence interpreting the estimated slope on the odds scale.
d. What is the estimated probability of survival for a 20-year-old?
e. What is the estimated relative "risk" of survival for a 60-year-old
compared to a 20-year-old.
f. Calculate and interpret a 95% confidence interval for the odds ratio
of survival for a 60-year-old compared to a 20-year-old.
Write a sentence interpreting this interval.

## Q3
Fit a logistic regression model to these data using only `sex` as the predictor.
Call this model `mod2`. What does this model estimate for the odds ratio
of survival for males compared to females?

## Q4
Create a 2x2 table with `sex` as the rows and `survive` as the columns.
Use this table to estimate the odds ratio of survival for males compared to females.
Does this value match your answer to part c.?

## Q5
Now, fit a logistic regression model using both `age` and `sex` as predictors,
with no interaction. Call this model `mod3`. Then answer the following:

a. Write out the equation of the fitted model for Males, and the equation
of the fitted model for Females.
b. Write a sentence interpreting the `sex` coefficient in this model.

## Q6
Lastly, fit a logistic regression model using both `age` and `sex` as predictors,
and include an interaction term. Call this model `mod4`.
Write an interpretation of each coefficient (except the intercept) in this model.

## Q7
Again, create a scatterplot with `age` on the x-axis and `survive` on the y-axis.
On the same plot, add the following:

- Fitted curve from `mod1`
- Fitted "curve" from `mod2`, one for each sex
- Fitted curve from `mod3`, one for each sex
- Fitted curve from `mod4`, one for each sex

Use different line types for the different models, and different colors
for the different sexes. Add a legend. Ensure your plot is well-labeled.

## Q8
Choose which model you think fits the data best. Then create an ROC curve
for your chosen model. How well does your model predict survival?



## Part II: Cite Sources

Write the sources you used to complete this assignment at the end of your homework submission, adhering to the "Guidance on Citing Sources" bullet points in the [collaboration policy section on our course syllabus](https://staceyhancock.github.io/stat439/index.html#Policy_on_collaboration_and_academic_misconduct).


