---
title: "GLMMs and GEEs with Binary Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4) # lmer
library(gee) # gee
library(vcd) # pairs
```

## Data on attitudes about abortion from Agresti (2013)

Subjects were asked whether they supported legalizing abortion in each of three situations: 

1. If the family has a very low income and cannot afford any more children.
2. When the woman is not married and does not want to marry the man.
3. When the woman wants it for any reason.

Gender was also measured.

Note: Since the model implies nonnegative association among responses,
situations and scales should correspond accordingly.
For example, with scale (yes, no), it would not make sense to ask
"Should abortion be legal...?" and " Should abortion be illegal...?"
as two different situations.

## Read in data

Variables:  
    * gender = 1 if female; 0 if male
    * response = 1 if support; 0 if do not support
    * question = situation (1, 2, or 3)
    * case = subject id

```{r}
attitudes <- read.table("https://math.montana.edu/shancock/data/abortion.txt", header=TRUE)
# Code situation as factor:
attitudes$question = factor(attitudes$question)
# Make situation 3 the reference situation:
attitudes$question = relevel(attitudes$question, ref="3")
# Note that we can leave gender and response as numeric
# since they are already coded as 0 or 1.
```


## Exploratory data analysis

Plot of proportion of support in each covariate pattern:
```{r}
means <- tapply(attitudes$response, list(attitudes$question,attitudes$gender), mean)
plot(c(3,1,2),means[1:3,1],type="p",col="red",pch=16,
	xlab="Situation",ylab="Proportion Support",
	main="Proportion Support by Situation and Gender",
	cex=2,ylim=c(0.45,.52))
points(c(3,1,2),means[1:3,2],col="darkblue",cex=2)	
legend("bottomleft", c("Male","Female"), col=c("red","darkblue"), pch=c(16,1))
```

Barplot of same information:
```{r base-R-barplot}
barplot(t(means), beside=TRUE, col=c("Red","Blue"),
	ylim=c(0,0.6), legend = c("Male","Female"),
	xlab="Situation",ylab="Proportion Support",
	main="Proportion Support by Situation and Gender")
```
```{r tidyverse-barplot}
# Convert data to long form - currently first col is male, second is female
means_long <- data.frame(Proportion = c(means[,1], means[,2]))
# Add gender and situation variables
means_long$Gender <- rep(c("Male", "Female"), each = 3)
means_long$Situation <- rep(c("Any reason", "Income", "Unmarried"), 2)

means_long %>% ggplot(aes(x = Situation, fill = Gender, y = Proportion)) +
  geom_col(position = "dodge")

barplot(t(means), beside=TRUE, col=c("Red","Blue"),
	ylim=c(0,0.6), legend = c("Male","Female"),
	xlab="Situation",ylab="Proportion Support",
	main="Proportion Support by Situation and Gender")
```

Contingency table summary of data by situation:
```{r}
my.tab <- xtabs(~response+gender+question, data=attitudes)
my.tab
```

Visualize three-way contingency table:
```{r}
pairs(my.tab, diag_panel = pairs_barplot,  labeling=TRUE)
```

Since we have binary data on three occasions, there are only 8 possible values
for the response vector.

```{r}
# Convert data to wide form
att_wide <- attitudes %>% pivot_wider(
  id_cols = "case",
  names_from = "question",
  names_prefix = "Sit",
  values_from = "response",
  unused_fn = list("gender" = max)
)
n <- dim(att_wide)[1]
```

The number of rows in wide form tell us the number of subjects questioned:
`r n`


How many supported legalizing abortion in all three situations, i.e., responded (1,1,1)?
```{r}
# Create sequence or responses variable:
# Hard way -
att_wide <- att_wide %>%
  mutate(response_seq = case_when(
    Sit1 == 1 & Sit2 == 1 & Sit3 == 1 ~ "111",
    Sit1 == 1 & Sit2 == 1 & Sit3 == 0 ~ "110",
    Sit1 == 1 & Sit2 == 0 & Sit3 == 1 ~ "101",
    Sit1 == 0 & Sit2 == 1 & Sit3 == 1 ~ "011",
    Sit1 == 1 & Sit2 == 0 & Sit3 == 0 ~ "100",
    Sit1 == 0 & Sit2 == 1 & Sit3 == 0 ~ "010",
    Sit1 == 0 & Sit2 == 0 & Sit3 == 1 ~ "001",
    Sit1 == 0 & Sit2 == 0 & Sit3 == 0 ~ "000",
    )
  )
# Easy way -
att_wide <- att_wide %>%
  mutate(response_seq = paste0(Sit1, Sit2, Sit3))

xtabs(~gender + response_seq, data = att_wide)
# Note that the majority of the sample was either 000 or 111.
# --> high association between repeated measurements.
# Sample correlations among repeated measurements:
cor(att_wide[,2:4])
```


## Modeling: GLMM

```{r}
### Fit GLMM by ML ###
# No interaction between gender and question -->
# gender effect assumed to be identical for each situation.
mod1 <- glmer(response ~ gender+question+(1|case), 
family=binomial,
	data=attitudes, nAGQ = 25)
# Larger value of "nAGQ" =  greater accuracy in approximation
# --> longer computational time	
# Default is 1 (Laplace approximation) - gives very different results!
summary(mod1)
summary(update(mod1, nAGQ = 1))
```


```{r}
mod2 <- glmer(response ~ gender*question+(1|case), family=binomial,
	data=attitudes, nAGQ = 25)
# Test for interaction effect between gender and question:	
anova(mod2)
anova(mod1,mod2)
```


```{r}
# EBLUPS same for all individuals with same response sequence
# and same gender, e.g.,
ranef(mod1)$case[att_wide$response_seq=="000", 1]
att_wide$gender[att_wide$response_seq=="000"]
```

```{r}
# Note different answers if we use a different R library:
library(glmmML) # May need to install package
# glmmML function only allows for a random intercept
mod1b = glmmML(response ~ gender+question, cluster=case,
	family=binomial, method="ghq", 
	n.points=70, start.sigma=9, data=attitudes)
summary(mod1b)	
```


## Modeling: GEE

```{r}
mod2 <- gee(response ~ gender+question, family=binomial,
            corstr="exchangeable", id=case, data=attitudes)
summary(mod2)	
```

Note since GEE is really an estimating procedure (similar to least squares) and there is no likelihood, we cannot carry out likelihood ratio tests to compare models.
