---
title: 'Poisson Regression: Horseshoe Crab Data'
date: '2022-03-08'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
```

# Poisson Regression: Horseshoe crab example

## Data set from Agresti (2007)

Study of nesting horseshoe crabs (J. Brockman, Ethology, 1996).
Each female crab in the study had a male crab attached to her in her nest.
The study investigated factors that affect whether the female crab had
any other males, called satellites, residing nearby her.
Explanatory variables thought possibly to affect this included the
female crab's color, spine condition, weight, and carapace width.
The response outcome for each female crab is her number of satellites.

**Variable descriptions:**

- `color`: 1 - light medium, 2 - medium, 3 - dark medium, 4 - dark
- `spine`: 1 - both good, 2 - one worn or broken, 3 - both worn or broken
- `width`: carapace width in cm
- `satell`: number of satellites
- `weight`: weight in kg

Read in the data and recode factors.
```{r}
crab <- read.table("http://www.math.montana.edu/shancock/data/horseshoe.txt",header=T)
crab$colorF <- factor(crab$color, levels=c(1,2,3,4), labels=c("LM","M","DM","D"))
crab$spineF <- factor(crab$spine, levels=c(1,2,3), labels=c("Good","Soso","Poor"))
str(crab)
```

## Exploratory data analysis

Summary statistics for each variable:
```{r}
summary(crab)
```

Plot all variables against one another:
```{r}
plot(crab)
```

Boxplots for categorical predictors:
```{r}
boxplot(satell ~ colorF, data=crab, xlab="Color", ylab="No. of Satellites")
boxplot(satell ~ spineF, data=crab, xlab="Spine Condition", ylab="No. of Satellites")
```

Based on the plots alone, what variables seem to be the best predictors of number of satellites?


## Poisson regression models

### One quantitative predictor: `width`

```{r}
crab.glm <- glm(satell ~ width, family = poisson(link = "log"), data = crab)
summary(crab.glm)
```

How would you interpret each coefficient?

Note that the log link is the default for poisson, so we could have used
`family = poisson` rather than `family = poisson(link = "log")`.

#### Checking model fit
How can we check if the model fits well?

Add fitted model to plot.
```{r}
co <- crab.glm$coefficients
plot(satell ~ width, xlab="Width (cm)", ylab="No. of Satellites", data=crab)
curve(exp(co[1]+co[2]*x),add=T)
```

#### Check for overdispersion

What is "overdisperson"?

Create the eight width categories used in book:

- 1 = <23.25
- 2 = 23.25-24.25
- 3 = 24.25-25.25
- 4 = 25.25-26.25
- 5 = 26.25-27.25
- 6 = 27.25-28.25
- 7 = 28.25-29.25
- 8 = >29.25 - Note: Book error = 30.25

One option to create width categories is the `cut` function.
Note that the resulting factor is treated as nominal even though it is ordinal.
```{r}
crab$width.grp <- cut(crab$width, breaks = c(0, seq(23.25,29.25,by=1), 35), labels=c(1:8))
is.factor(crab$width.grp)
head(cbind(crab$width, crab$width.grp))
```

Explore association between number of satellites and width group.
```{r}
xtabs(~ satell+width.grp, data=crab)
```

Since `width.grp` is a factor, R defaults to side-by-side boxplots when we use `plot`.
```{r}
plot(satell ~ width.grp, data=crab,
     xlab = "Width Group", ylab = "Number of Satellites")
```

Examine our fitted model over the means for each group.

```{r}
means <- tapply(crab$satell, crab$width.grp, mean)
plot(seq(22.75,29.75,by=1), means, 
     xlab="Width", ylab="Mean No. of Satellites")
curve(exp(co[1]+co[2]*x), add=T)
```

Check for overdispersion by comparing the means within each group to
the sample variances.
```{r}
vars <- tapply(crab$satell, crab$width.grp, var)
kable(data.frame(width.grp = 1:8, means = round(means, 3), variances = round(vars,3)))
```
The variances in each group are much larger than the means for each group,
indicating overdispersion is present.
This may indicate need to include other predictors to reduce unexplained variability.


### One categorical predictor: `colorF`

```{r}
crab.glmF <- glm(satell ~ colorF, family = poisson(link = "log"), data = crab)
summary(crab.glmF)
```

How would you interpret each coefficient?


### Poisson regression with two predictors: `width` and `colorF`

Model with interaction:
```{r}
crab.glm2 <- glm(satell ~ width*colorF, data=crab, family=poisson)
summary(crab.glm2)
```

How to interpret coefficient estimates?

Plot number of satellites vs width by color and add fitted model.
```{r}
co <- crab.glm2$coef
plot(satell ~ width, data=crab, xlab="Width (cm)", ylab="No. of Satellites", 
     pch=color, col=(color+1))
legend("topright", c("LM","M","DM","D"), pch=c(1,2,3,4), col=c(2,3,4,5))
# Add fitted models by color
# LM
curve(exp(co[1]+co[2]*x), col=2, add=TRUE)
# M
curve(exp(co[1]+co[3] + (co[2]+co[6])*x), col=3, add=TRUE)
# DM
curve(exp(co[1]+co[4] + (co[2]+co[7])*x), col=4, add=TRUE)
# D
curve(exp(co[1]+co[5] + (co[2]+co[8])*x), col=5, add=TRUE)
```


Or with the effects package:
```{r}
library(effects)
plot(allEffects(crab.glm2), type="response", multiline=T, ci.style="bands")
plot(allEffects(crab.glm2), type="link", multiline=T, ci.style="bands")
plot(allEffects(crab.glm2, residuals=T), type="link")
```


Additive model (no interaction):
```{r warning=F, message=F, fig.height=7, fig.width=10}
crab.glm2a <- glm(satell ~ width + colorF, data=crab, family=poisson)
co <- crab.glm2a$coef
plot(satell ~ width, data=crab, xlab="Width (cm)", ylab="No. of Satellites", pch=color, col=(color+1))
legend("topright", c("LM","M","DM","D"), pch=c(1,2,3,4), col=c(2,3,4,5))
# LM
curve(exp(co[1]+co[2]*x), col=2, add=TRUE, lty=2)
# M
curve(exp(co[1]+co[3] + (co[2])*x), col=3, add=TRUE, lty=2)
# DM
curve(exp(co[1]+co[4] + (co[2])*x), col=4, add=TRUE, lty=2)
# D
curve(exp(co[1]+co[5] + (co[2])*x), col=5, add=TRUE, lty=2)
```


What if we treated color as quantitative? (Is this valid?)
```{r warning=F, message=F, fig.height=7, fig.width=10}
crab.glm3 <- glm(satell ~ width*color, data=crab, family=poisson)
summary(crab.glm3)
# Note significant interaction term.
```

```{r}
plot(satell ~ width, data=crab, xlab="Width (cm)", ylab="No. of Satellites", pch=color, col=(color+1))
legend("topleft", c("LM","M","DM","D"), pch=c(1,2,3,4), col=c(2,3,4,5))
## Add fitted model where color is quantitative to plot:
co <- crab.glm3$coef
# LM
curve(exp(co[1] + co[2]*x + co[3]*1 + co[4]*x*1), col=2, add=TRUE, lty=3, lwd=2)
# M
curve(exp(co[1] + co[2]*x + co[3]*2 + co[4]*x*2), col=3, add=TRUE, lty=3, lwd=2)
# DM
curve(exp(co[1] + co[2]*x + co[3]*3 + co[4]*x*3), col=4, add=TRUE, lty=3, lwd=2)
# D
curve(exp(co[1] + co[2]*x + co[3]*4 + co[4]*x*4), col=5, add=TRUE, lty=3, lwd=2)
```

```{r}
# Effect of a one cm increase in width for each color:
for(i in 1:4){
	print(exp(crab.glm3$coef[2]+crab.glm3$coef[4]*i))
}
```

